---
title: "Codebook"
author: "Ben Graham"
date: "`r Sys.Date()`"
output: html_document
---

## The Project

For this project, we transformed a raw dataset into a tidy dataset, preparing it
for future work. Please refer to the README.Rmd in this repository for details 
on the original dataset. This project takes a raw dataset and transforms it 
into tidy data suitable for further analysis.

##The script: run_analysis.R

The script breaks the project into 7 parts, and it depends on the dplyr and 
data.table packages to run. 

#1. Download & Unzip into working directory

```{r}
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"  
dataset <- c("dataset.zip")
download.file(url,destfile=dataset,mode='wb')
unzip(zipfile="./dataset.zip")
```

#2. Import 8 required files into R. 

```{r}
# Descriptions
features <- read.table("UCI HAR Dataset/features.txt",col_names=c("n","measure"))
activities <- read.table("UCI HAR Dataset/activity_labels.txt",col_names=c("code","activity"))

# Training Data
subject_train <- read.table("UCI HAR Dataset/train/subject_train.txt", col.names = "subject")
x_train <- read.table("UCI HAR Dataset/train/X_train.txt", col.names = features$measure)
y_train <- read.table("UCI HAR Dataset/train/y_train.txt", col.names = "code")

# Test Data
subject_test <- read.table("UCI HAR Dataset/test/subject_test.txt", col.names = "subject")
x_test <- read.table("UCI HAR Dataset/test/X_test.txt" , col.names = features$measure)
y_test <- read.table("UCI HAR Dataset/test/y_test.txt" , col.names = "code")
```

The column names for the x_test and x_train data are in the features.txt file, 
and are imported as shown. 

#3. Combine datasets

Six individual tables from the training and test data will be combined to form
our final dataset, as shown.

  A tibble: 2 x 3
  subject       y_data  x_data 
  <chr>         <chr>   <chr>  
1 subject_train y_train x_train
2 subject_test  y_test  x_test 

First, we'll column bind our training and test sets together horizontally, and  
then we'll row bind the two sets together vertically. 

```{r}
# Bind Columns for Training & Test Sets
data_train <- bind_cols(subject_train,y_train,x_train)
data_test <- bind_cols(subject_test,y_test,x_test)

# Bind Training and Test Sets
data_all <- bind_rows(data_train,data_test)
```

#4. Subset data to include only mean & standard deviation values

Next, we'll subset our data to include only our subject and code variables, and
any columns that include mean or standard deviation data. I made a copy of the 
column names at this point, before we continue. 

```{r}
# Create data subset
data_tidy <- data_all %>% select (subject,code,contains("mean"),contains("std"))
# Create vector of column names
untidy_names <- colnames(data_tidy)
```

#5. Transform Column names into descriptive titles 

To transform the column names into descriptive titles, we use the gsub function
to change multiple names at once. Example as shown:

```{r}
names(data_tidy) <- gsub(".mean...X"," Mean(X)",names(data_tidy))
```

I probably could have done this much more efficiently, but once I was most of
the way through, I didn't want to start over. At this point, I saved a copy of 
the new tidy column names and combined them with the untidy names in a tibble
as a quick reference. 

```{r}
# Create copy of new column names
tidy_names <- colnames(data_tidy)
# Reference of old names / new names
name_changes <- tibble(untidy_names,tidy_names)
```

#6. Create final summary table of data

Next, we created a summary table that shows the average value for each activity,
for each participant in the study, and swap out the code column for the activity 
entry to make it more legible. 

```{r}
# Create final summary table
data_final <- data_tidy %>%
        group_by(subject,code) %>%
        summarize_all(mean)
        
# Swap Code for Activity description
data_final$code <- activities[match(data_final$code,activities$code), ]$activity
data_final <- data_final %>% rename(Activity=code)
```

#7. Write a copy of the new tidy dataset to the working directory.

Finally, we write a copy of the tidy dataset to Tidy.txt for inclusion in the 
repository.

```{r}
# Write txt file from summary
write.table(data_final, file = "Tidy.txt", row.names = FALSE)
```

#8. Upload to Github

Final upload to Github includes this Codebook, a README file, the run_analysis.R
script, and the Tidy.txt cleaned dataset. 





